---
output: html_fragment
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(tomhes)
library(dplyr, warn.conflicts = FALSE, quietly = TRUE)
library(ggplot2)
library(RcappeR)
library(magrittr)
library(reshape2)
```

```{r echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE}
## This code unfortunately can't be run by others.
mongo <- mongo.create()
count <- mongo.count(mongo = mongo, ns = "gulf.ratings")
cursor <- mongo.find(mongo = mongo, ns = "gulf.ratings")
output <- vector("list", count)
i <- 1
while(mongo.cursor.next(cursor)) {
    output[[i]] <- mongo.bson.to.list(mongo.cursor.value(cursor))
    i <- i + 1
}

df <- gulf_ratings_df(ratings = output)
df %<>%
    group_by(date_race) %>%
    arrange(pos) %>%
    ungroup()

buf <- mongo.bson.buffer.create()
mongo.bson.buffer.append(buf = buf, name = "date", value = "31/01/15")
query <- mongo.bson.from.buffer(buf = buf)
count <- mongo.count(mongo = mongo, ns = "gulf.races", query = query)
cursor <- mongo.find(mongo = mongo, ns = "gulf.races", query = query)
sat_races <- vector("list", count)
i <- 1
while(mongo.cursor.next(cursor)) {
    sat_races[[i]] <- mongo.bson.to.list(mongo.cursor.value(cursor))
    i <- i + 1
}
sat_races <- gulf_races_df(races = sat_races)
sat_race <- sat_races %>%
    filter(race == 8) %>%
    select(date_race, date, race, race_type, dist, surf, value, race_cond,
           clm_price, f_m, runup, going, trainer, jockey, j_clm, age, wgt, gate,
           pos, horse, fintime,
           c2f, c3f, c3_5f, c4f, c4_5f, c5f,
           fs3f, fs2f, fs1_5f, fs1f, fs1_5f,
           d2f, d3f, d3_5f, d4f, d4_5f, d5f,
           ss2f, ss3f, ss3_5f, ss4f, ss4_5f, ss5f)
```

## Saturday 31/1/15

This first blog, will look to handicap (calculate a rating) for a race run at Gulfstream Park over the weekend.  The race in question is Race #`r unique(sat_race$race)`, an `r unique(sat_race$race_type)` race run over `r unique(sat_race$dist)` furlongs on the `r unique(sat_race$surf)`.  A quick look at the result:  

```{r echo=FALSE}
sat_race %>%
    select(wgt, pos, horse, fintime) %>%
    arrange(pos)
```

_I use Trakus for final times, which recorded this race as a dead heat.  The official result, I believe, recorded `r subset(sat_race, pos == 1)$horse` as a winner by a nose, the difference between a dead heat and a nose is about two 100ths of a second._

Ahead of handicapping the race, I'll have a look at the race first using the sectional speeds of horses, these are calculated using Trakus' distance raced and sectional data.  Sectional speeds are not cumulative, so it charts the speed from one sectional to the next, for a 5 furlong race, these intervals are from 0 to 2f, 2f to 3f, and from 3f to the finish intervals are 0.5f.  The plot below shows the sectional speeds of the first four.  (The legend isn't too clear using the custom ggplot2 theme included in the `RcappeR` package, this will be fixed.)

```{r echo=FALSE, warning=FALSE, message=FALSE, out.height="340px", out.width="550px", fig.align='center'}
sat_race_speeds <- sat_race %>% select(pos, horse, starts_with("ss"))
sat_race_speeds <- melt(sat_race_speeds, id.vars = c("pos", "horse"), value.name = "speed", variable.name = "sectional")

sat_race_speeds$sectional %<>% gsub(pattern = "ss", replacement = "", x = .) %>%
    gsub(pattern = "_", replacement = ".", x = .) %>%
    gsub(pattern = "f", replacement = "", x = .) %>%
    as.numeric()

sat_race_speeds$horse <- factor(sat_race_speeds$horse, levels = unique(sat_race_speeds$horse))

ggplot(subset(sat_race_speeds, pos <= 4), aes(x = sectional, y = speed, col = horse)) +
    geom_smooth(group = 1, size = 1, alpha = .75) +
    theme_rcapper()
```

The winner, `r subset(sat_race, pos == 1)$horse`, and the fourth placed, `r subset(sat_race, pos == 4)$horse` ran remarkably similar races, the difference was in the opening 2f, where the winner opened up a `r subset(sat_race, pos == 4)$c2f - subset(sat_race, pos == 1)$c2f` second advantage, the difference at the finish was `r subset(sat_race, pos == 4)$fintime - subset(sat_race, pos == 1)$fintime`.  The difference between the winner and the second place is more apparent, with `r subset(sat_race, pos == 2)$horse` running the opening 2f in a speed of `r subset(sat_race, pos == 2)$ss2f`, as a result, tiring towards the finish.

Finishing Speeds can reveal how efficiently a horse ran its race, they are another innovation by Simon Rowlands, see parts [1](https://www.timeform.com/Racing/Articles/An_introduction_to_Sectional_Timing_Part_1) and [2](https://www.timeform.com/Racing/Articles/An_introduction_to_Sectional_Timing_Part_2) for an introduction.  It expresses the finishing sectional as a percentage of average race speed, so above 1, means the finishing sectional was faster than average race speed.

The par Finishing Speed over the final 2f for 5f turf races is **`r df %>% filter(dist == 5, pos == 1) %>% summarise(fs = mean(fs2f)) %>% unlist() %>% as.vector() %>% round(3)`**, and while the plot above makes it look like `r subset(sat_race, pos == 2)$horse` was slowing, her Finishing Speed of `r subset(sat_race, pos == 2)$fs2f` is not that different to the par.  She ran a pretty efficient race, on the other hand, the eventual winner, `r subset(sat_race, pos == 1)$horse`, who got up in the dying strides, recorded a Finishing Speed of `r subset(sat_race, pos == 1)$fs2f`, well above par.

Simon converts Finishing Speeds to a percentage, so `r subset(sat_race, pos == 2)$fs2f` becomes `r subset(sat_race, pos == 2)$fs2f * 100`, the `fin_spd` function in `RcappeR` follows Simon, however Finishing Speeds found on this site are left as decimals, to help distinguish quickly between a final time and a Finishing Speed on an individual horses page.

For more about the 5f turf course, see the [track profile](http://durtal.github.io/gulfstream/#/track-profiles/5-furlong-profile).

### Handicapping

As explained (potentially poorly) in the handicapping method post found in the ratings section, I calculate performance ratings using similar races and the performances of runners from those similar races.  So `r subset(sat_race, pos == 1)$horse` will be assessed using other runners in `r unique(sat_race$race_type)` races.  These posts will hopefully crystalise the method explained in the lengthy handicapping method post.  

The performance ratings can be improved upon, some improvements will be quite simple (a more flexible lbs per second scale for a start), while others are more complicated and may take me some time.  (Any questions, suggestions, queries, please ask, my twitter account and email can be found on this site, and I'll do my best to answer.)

The plot below shows the distribution of ratings for the winners of similar races:

```{r echo=FALSE, out.height="340px", out.width="500px", fig.align='center'}
COR <- unique(paste("all op clm / all", stringr::str_extract(string = sat_race$race_cond, pattern = "nw\\d"), sep = " - "))
subsetraces <- df %>%
    filter(cor == COR) %>%
    group_by(date_race) %>%
    arrange(pos) %>%
    ungroup()

subsetraces %>%
    filter(pos == 1) %>%
    ggplot(aes(x = RRtg, y = ..density..)) +
        geom_bar(binwidth = 1, fill = "#D9220F", col = "#FCFCFC") +
        theme_rcapper()
```

#### Data Preparation

Ahead of calculating a rating for the runners in Race #`r unique(sat_race$race)` on Saturday, the data needs to be prepped.  The [handicapping preparation](http://durtal.github.io/RcappeR/data_prep.html) vignette included in my [RcappeR](https://github.com/durtal/RcappeR) package walks through an example, but I'll walk through this race as well.  Data is already clean, in a workable format, and the variables needed are present, seen above.

I'll be using dplyr and magrittr to manipulate my dataset, below are the steps, each includes a bit of code that is then used in the following code section:

* arrange dataset by finishing position
    + `sat_race %<>% arrange(pos) %>%`
* calculate new variables using `mutate` from `dplyr`
    + calculate beaten seconds (uses `btn_sec` from `RcappeR`)
        + `btn_sec = btn_sec(times = fintime)`
    + calculate lbs per second scale (uses `lbs_per_sec` from `RcappeR`)
        + `scale = lbs_per_sec(dist = dist)`
    + calculate beaten lbs
        + `btn_lbs = scale * btn_sec`
    + calculate a difference at the weights (uses `diff_at_wgts` from `RcappeR`)
        + `diff_at_wgts = diff_at_wgts(btn_lbs = btn_lbs, wgt_carried = wgt)`

```{r}
sat_race %<>%
    arrange(pos) %>%
    mutate(btn_sec = btn_sec(times = fintime),
           scale = lbs_per_sec(dist = dist),
           btn_lbs = scale * btn_sec,
           diff_at_wgts = diff_at_wgts(btn_lbs = btn_lbs, wgt_carried = wgt))

sat_race %>%
    select(wgt, pos, horse, fintime, btn_sec, scale, btn_lbs, diff_at_wgts)
```

#### Rating

With the dataset prepped, the larger dataset containing many similar races, that have already been handicapped and ratings calculated, can be used to calculate a rating for `r subset(sat_race, pos == 1)$horse`.  This process uses the `zipf_hcp` function from `RcappeR`.  

In short, for each race in the larger dataset, the margins (the `diff_at_wgts` variable) between horses in `r subset(sat_race, pos == 1)$horse`'s race are added, according to finishing position, to the ratings of runners in the race from the larger dataset, resulting in a vector of ratings.  This vector represents winning ratings according to finishing position, these are then weighted inversely to rank (position), so the rating in position 2 in the vector is multiplied by 1/2, position 3 by 1/3, etc (this is the use of Zipf's Law).  These weighted ratings are then summed and divided by the sum of the weights (1/1 + 1/2 + 1/3, etc), resulting in a single rating.  When many races from the larger dataset are used a distribution of potential ratings is built up.

(For a more detailed explanation of the process see the [zipf_race](http://durtal.github.io/RcappeR/handicap_with_zipf_race.html) vignette.)

The `zipf_hcp` function comes with a few S3 methods, perhaps the most useful (pretty), is the plot method, which shows the distribution of ratings, the mean rating, and the number of ratings used.

```{r fig.align='center', out.height="340px", out.width="500px"}
hcp <- zipf_hcp(race = sat_race, past_races = subsetraces, race_id = "date_race", btn_var = "diff_at_wgts", rating = "RRtg")

plot(hcp)
```

So using `r hcp$n_races` similar races, `r subset(sat_race, pos == 1)$horse` is awarded a rating of **`r hcp$mean_rt`**.

The ratings for all runners in the race are as follows:

```{r echo=FALSE}
sat_race %<>%
    mutate(RRtg = hcp$mean_rtg - diff_at_wgts)

sat_race %>%
    select(wgt, pos, horse, fintime, btn_lbs, diff_at_wgts, RRtg) %>%
    arrange(pos)
```

Only horses who have raced at least 5 times at Gulfstream Park since 2012 are included in the app at the moment, this means the winner isn't present at the moment, the second horse is, see her race record [here](http://durtal.github.io/gulfstream/#/ratings/horse/spectacular_me_2010) (hopefully that link works!)

### Improvements

There are ways that these ratings can be improved, I'll quickly rifle through a few of them.

At the moment ratings don't take into account the ability of runners in the individual races.  I am somewhat limited by only having data for results from Gulfstream Park, so a horses entire race history is unknown to me (there are a lot of horses who have raced just once).  This also hinders me from calculating master ratings for horses (though this is not sole reason).

The rigid lbs per second scale greatly inflates wide margin winners.  This impacts on Maiden races especially, where runners are at various stages of maturity, readyness and ability.  A more flexible lbs per second function will take into account the speed of individual runners in a given race.  So each horse gets its own lbs per second scale.

Speed ratings.  Now I have some basic ratings, I can hopefully look to calculate standard times, from which performances of runners can be assessed against the clock as well as against it's rivals in a race.  Helping to build a more complete picture of a horses ability.

Finally a few things I can do now, **READ MORE**, **LEARN MORE**.
