---
output: html_fragment
---

## Data

This app contains data for horses to have raced at Gulfsteam Park since December 2012, the data can be found in the github repo for this app, find the repo [here](https://github.com/durtal/gulfstream).  The data is [JSON](http://en.wikipedia.org/wiki/JSON) and this post will walk through loading the JSON into [R](http://cran.r-project.org/), and converting it into a neat dataframe.  A subset of this data can also be found in my [RcappeR](https://github.com/durtal/RcappeR) package.

### Load libraries

First we'll need a couple of libraries, one to convert the JSON into a list (`jsonlite`), and the second (`plyr`) to help with turning the list to a dataframe.  We'll also need a couple of helper functions found in this [gist](https://gist.github.com/durtal/f0ca54e7aacb92130a1e), the functions are `gulf_ratings` and `gulf_ratings_helper`, the former calls the latter.

```{r message=FALSE, warning=FALSE}
library(jsonlite)
library(plyr)

# we'll use the source_gist function from devtools so
# install.packages("devtools")
devtools::source_gist("https://gist.github.com/durtal/f0ca54e7aacb92130a1e")
```

### Read JSON into R

The `fromJSON` function from `jsonlite` is used, it will attempt to convert the data to a dataframe, but as the JSON is nested, an example of a single horse's data will follow, we need to tell `fromJSON` not to simplify to a dataframe.

```{r message=FALSE, warning=FALSE, cache=TRUE}
ratings_json <- fromJSON(txt = "https://raw.githubusercontent.com/durtal/gulfstream/master/app/data/gulfstream-ratings.json", simplifyDataFrame = FALSE)
```

We should now have a large list in our workspace, at the time of writing (`r Sys.Date()`) the list has
**`r length(ratings_json)`**.

### A single horse

We can view what the JSON looks like, each horse will have the following structure.

```{r, echo=FALSE}
horse <- toJSON(ratings_json[[1]][c("_id", "yob", "horse", "horse_abr", "trainer", "n_races")], pretty = TRUE)
cat(horse)
```

Each horses will also contain a races object, that will contain the individual races (the number of races is stored in the JSON as `n_races`, see above) they competed in, only one race is shown here.  For a full list of variables see the bottom of the post.

```{r echo=FALSE}
a_race <- toJSON(ratings_json[[1]]$races[1], pretty = TRUE)
cat(a_race)
```

### Convert to Dataframe

Thanks to the helper functions the list can be converted to a dataframe easily, due to the nested nature it can take some, `gulf_ratings` uses `plyr`'s `ldply` function, and it also includes `plyr`'s progress bar (which is not used here).

```{r, cache=TRUE}
ratings_df <- gulf_ratings(ratings = ratings_json, .progress = "none")
```

We should now have a nice dataframe, with `r length(names(ratings_df))` variables.  A look at the first few rows, shows that the dataframe is arranged by horse, and then by date (most recent races at the top):

```{r}
head(ratings_df)
```

When this post was written the app contained data for horses who have raced at Gulfstream Park at least 5 times since December 2012, this means that races in the dataset might not have all runners from that race.  Each race in the dataset can be identified using the `date_race` variable

### Variables

There are 29 variables in the dataset, the majority should be self explanatory, but nevertheless definitions can be found here.

- **id** unique horse id, consisting of horses' name and year of birth (used as slug in app).
- **yob** year of birth
- **current_trainer** trainer of horse in its last race
- **.id** id number for race in a horses race record
- **date** date of race
- **race** race number (on race card)
- **date_race** serves as more accurate race id
- **race_type** type of race: Claiming == "clm", Maiden Claiming == "mdn clm", etc
- **dist** race distance (furlongs)
- **surf** race surface (dirt or turf)
- **value** race value (or purse)
- **race_cond** any additional conditions on its runners (Non Winners 1 == "NW1")
- **clm_price** maximum claiming price (if applicable)
- **f_m** filly and mares only race
- **runup** length of run-up before clock starts (in feet)
- **going** official going
- **cor** class of race (I use this to handicap, combination of race_type and various other conditions)
- **trainer** trainer for the race
- **jockey** jockey
- **j_clm** jockey claim (in lbs)
- **age** horse age
- **wgt** weight carried (in lbs)
- **gate** starting gate
- **pos** finishing position
- **horse** horse
- **fintime** finishing time
- **fs2_5f** 2.5 furlong Finishing Speed
- **fs2f** 2 furlong Finishing Speed
- **RRtg** my RcappeR rating for the horses performance