---
output: html_fragment
---

## Race Data

As mentioned in an earlier post (and elsewhere in the app), the app contains data for horses who have raced 5 time or more at Gulfstream Park since December 2012.  A recent addition to the app made individual races available, and thereby including every runner I have rated, you can read about that update [here](http://durtal.github.io/gulfstream/#/blog/Update-27-04-15).  This means, that a more complete dataset can be built rather than the incomplete dataset collected in an older [post](http://durtal.github.io/gulfstream/#/blog/gulfstream-data).

This post will walk through loading the JSON into [R](http://www.r-project.org/), and converting it into a neat dataframe.  A subset of this data can also be found in my [RcappeR](https://github.com/durtal/RcappeR) package.

### Load libraries

We need a couple of libraries, one to convert the JSON into a list (`jsonlite`) and a second (`plyr`) to help with turning the list into a dataframe.  We'll also need a couple of helper functions found in this [gist](https://gist.github.com/durtal/f4c02bda609bb79b59d0), the helper functions are `gulf_races` and `gulf_races_helper`, the former calls the latter.

```{r message=FALSE, warning=FALSE}
library(jsonlite)
library(plyr)

# we'll use the source_gist function from devtools so
# install.packages("devtools")
devtools::source_gist("https://gist.github.com/durtal/f4c02bda609bb79b59d0")
```

### Read JSON into R

The `fromJSON` function from `jsonlite` is used, it will attempt to convert the data to a dataframe, but as the JSON contains some nested elements, an example of a single race object will follow, we need to tell `from JSON` not to simplify to a dataframe.

```{r message=FALSE, warning=FALSE, cache=TRUE}
races_json <- fromJSON(txt = "https://raw.githubusercontent.com/durtal/gulfstream/master/app/data/gulfstream-races.json", simplifyDataFrame = FALSE)
```

We should now have a large list in our workspace, at the time of writing (`r Sys.Date()`) the list has **`r length(races_json)`** unique races.

### A Single Race

We can view what the JSON looks like, each race will have the following structure.

```{r echo=FALSE}
race <- (toJSON(races_json[[1]][c("date", "date_race", "race", "details", "conditions")], pretty = TRUE))
cat(race)
```

Each race object will contain a runners object, which will contain data on the individual runners in a race, each runner has the following variables:

```{r echo=FALSE}
names(races_json[[1]]$runners[[1]])
```

### Convert to Dataframe

Thanks to the helper functions the list can be converted to a dataframe easily, due to the nested structure it can take some time.  `gulf_races` uses `plyr::ldply` function and includes the progress bar as a function argument (which is not used here, but change "none" to "text" or "time").

```{r cache=TRUE}
races <- gulf_races(races = races_json, .progress = "none")
```

This should return a dataframe with `r length(names(races))` variables, a few of the variables are in character format, due to `jsonlite` I believe, but should really be in numeric format, an example is the _fintime_ variable.  A look at the first few rows:

```{r}
head(races)
```

```{r echo=FALSE}
dr1 <- head(races$date, 1)
dr2 <- head(races$race, 1)
race_u <- paste0("http://durtal.github.io/gulfstream/#/ratings/date/", dr1, "/race/", dr2)
```
The variable _date\_race_ serves as a unique race id, it can be seen in the above rows, that the rows contain runners from the same race, run on **`r head(races$date, 1)`** in race number **`r head(races$race, 1)`** and the runners are ordering according to the _pos_ (position) variable.  The race can be seen in the app [here](`r race_u`).
