---
output: html_fragment
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
## This code unfortunately can't be run by others.
library(tomhes)
library(dplyr, warn.conflicts = FALSE, quietly = TRUE)
library(ggplot2)
library(reshape)

dist = 5.5

mongo <- mongo.create()
if(mongo.is.connected(mongo)) {
    buf <- mongo.bson.buffer.create()
    mongo.bson.buffer.append(buf, name = "details.dist", value = dist)
    query <- mongo.bson.from.buffer(buf = buf)
    count <- mongo.count(mongo = mongo, ns = "gulf.races", query = query)
    cursor <- mongo.find(mongo = mongo, ns = "gulf.races", query = query)
    output <- vector("list", count)
    i <- 1
    while(mongo.cursor.next(cursor)) {
        output[[i]] <- mongo.bson.to.list(mongo.cursor.value(cursor))
        i <- i + 1
    }
    df <- gulf_races_df(races = output)
}
rm(buf, count, cursor, i, output, query)
fill = "#D9220F"
```
_Last Updated `r Sys.Date()`_

# Dirt

This is the track profile for all `r dist` furlong `r unique(df$surf)` races at Gulfstream Park, there are **`r length(unique(df$date_race))`** races in the dataset, contested by **`r length(df$date)`** runners.  The oldest race in the dataset was run on **`r min(as.Date(df$date, format = "%d/%m/%y"))`** while the most recent was run on **`r max(as.Date(df$date, format = "%d/%m/%y"))`**.  First a few summaries about the types of races.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
mongo.destroy(mongo)
rm(list=ls())
```