---
output: html_fragment
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
## This code unfortunately can't be run by others.
library(tomhes)
library(dplyr, warn.conflicts = FALSE, quietly = TRUE)
library(ggplot2)
library(reshape)

dist <- 6

mongo <- mongo.create()
if(mongo.is.connected(mongo)) {
    buf <- mongo.bson.buffer.create()
    mongo.bson.buffer.append(buf, name = "details.dist", value = dist)
    query <- mongo.bson.from.buffer(buf = buf)
    count <- mongo.count(mongo = mongo, ns = "gulf.races", query = query)
    cursor <- mongo.find(mongo = mongo, ns = "gulf.races", query = query)
    output <- vector("list", count)
    i <- 1
    while(mongo.cursor.next(cursor)) {
        output[[i]] <- mongo.bson.to.list(mongo.cursor.value(cursor))
        i <- i + 1
    }
    df <- gulf_races_df(races = output)
}
rm(buf, count, cursor, i, output, query)
fill = "#D9220F"
wnrs <- filter(df, pos == 1)
```
_Last Updated `r Sys.Date()`_

# Dirt

This is the track profile for all `r dist` furlong dirt races at Gulfstream Park, there are **`r length(unique(df$date_race))`** races in the dataset, contested by **`r length(df$date)`** runners.  The oldest race in the dataset was run on **`r min(as.Date(df$date, format = "%d/%m/%y"))`** while the most recent was run on **`r max(as.Date(df$date, format = "%d/%m/%y"))`**.

### Summary

```{r echo=FALSE, warning=FALSE, message=FALSE}
racetypes <- df %>%
    filter(pos == 1) %>%
    group_by(race_type) %>%
    summarise(n = n()) %>%
    arrange(desc(n))

racetypes$race_type <- factor(racetypes$race_type, levels = racetypes$race_type)

fieldsize <- df %>%
    group_by(date_race) %>%
    summarise(field = n()) %>%
    group_by(field) %>%
    summarise(count = n())

fieldsize$n <- fieldsize$field * fieldsize$count
fieldsize$field <- factor(fieldsize$field, levels = fieldsize$field)
```

A few quick and dirty summary plots highlighting the different race types and field sizes.  **`r toupper(paste(racetypes$race_type[which.max(racetypes$n)]))`** races are the most common over the `r dist` furlong course, with **`r racetypes$n[which.max(racetypes$n)]`** races.  The average field size for `r dist` furlong races is **`r round(sum(fieldsize$n) / sum(fieldsize$count), 2)`**.

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.show='hold', out.height="300px", out.width="400px"}
ggplot(racetypes, aes(x = race_type, y = n)) +
    geom_bar(fill = fill, stat = "identity") +
    RcappeR::theme_rcapper() +
    labs(x = "Race Type", y = "Count", title = "Counts per Race Type (6f dirt)") +
    theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1, size = rel(1)))

ggplot(fieldsize, aes(x = field, y = count)) +
    geom_bar(fill = fill, stat = "identity") +
    RcappeR::theme_rcapper() +
    labs(x = "No. of Runners", y = "Count", title = "Field Size (6f dirt)")

rm(racetypes, fieldsize)
```

### Draw

The six furlong course is run round one turn, which starts roughly 2 furlongs from the starting gates, and continues for roughly 2 furlongs, meaning the home straight is ~2f.  Counting winners is the typical way to assess any draw bias, but winners ignore a large number of runners, and if there was any such bias, it should be seen across all runners, so percentage of runners beaten is to be used (gates must have a minimum of 5 runs):

```{r echo=FALSE, fig.align="center", out.height="350px", out.width="500px"}
tmp <- df %>%
    group_by(gate) %>%
    mutate(rnrs = n()) %>%
    ungroup() %>%
    filter(rnrs >= 5) %>%
    group_by(date_race) %>%
    mutate(rnrs_btn = RcappeR::rnrs_btn(pos)) %>%
    ungroup() %>%
    group_by(gate) %>%
    summarise(pct_btn = mean(rnrs_btn, na.rm = TRUE),
              mean_pos = mean(pos, na.rm = T))

tmp$gate <- factor(tmp$gate, levels=tmp$gate)
ggplot(data = tmp, aes(x = gate, y = pct_btn)) +
    geom_bar(fill = fill, stat = "identity") +
    RcappeR::theme_rcapper() +
    labs(title = "Percentage of Rivals Beaten (6f dirt)")

morethan55 <- tmp$gate[which(tmp$pct_btn >= .55)]
```

A fair draw would expect to see each draw beat 50% of their rivals`r if(length(morethan55) > 0) { paste(", runners draw in", if(length(morethan55) >= 2) { paste("gates") } else { paste("gate") }) } else { paste(".") }` `r if(length(morethan55) > 0) { paste(morethan55, collapse = ", ") }` `r if(length(morethan55) > 0) { paste(" beat more than 55% of their rivals.") }`

```{r echo=FALSE}
tmp <- df %>%
    group_by(gate) %>%
    summarise(n = n(),
              mean_dist = mean(d6f, na.rm = TRUE))

tmp <- filter(tmp, n >= 5)

lowgate <- tmp$gate[which.min(tmp$mean_dist)]
lowdist <- round(subset(tmp, gate == lowgate)$mean_dist, 2)
highgate <- tmp$gate[which.max(tmp$mean_dist)]
highdist <- round(subset(tmp, gate == highgate)$mean_dist, 2)
```

The draw will likely have an impact on the distance horses travel, especially as the first and only turn is such a short distance from the start.  Horses drawn wide (high) will likely be forced wide.  `r if(tmp$gate[which.min(tmp$mean_dist)] == 1) { paste("Unsurprisingly, horses") } else { paste("Horses") }` drawn in gate **`r lowgate`** travel the least distance, with an average trip of **`r lowdist`** feet, while horses draw in gate **`r highgate`** travel an average distance of **`r highdist`**, a difference of **`r highdist - lowdist`** feet (gates must have a minimum of 5 runs used).  The plot below shows boxplots of distance raced by horses in each starting gate.

```{r, fig.align="center", out.height="350px", out.width="500px", echo=FALSE, warning=FALSE, message=FALSE}
tmp <- df %>%
    group_by(gate) %>%
    mutate(runs = n()) %>%
    ungroup() %>%
    filter(runs >= 5)

ggplot(tmp, aes(x = factor(gate), y = d6f)) +
    geom_boxplot(fill = fill) +
    RcappeR::theme_rcapper() +
    labs(x = "Gate", y = "Distance (ft)", title = "Distance Raced from each Starting Gate (6f dirt)")

rm(tmp, lowdist, lowgate, highdist, highgate)
```

### Race Shape

Race shape is something that is far easier to highlight at sprint distances, they are the fastest run races and as such take on a similar shape.  That is, by and large, all runners are slowing at the finish, almost irrespective of the pace they showed early in the race - of course the extent of the slowing will be down to pace.  The plot below shows the sectional speeds of winners, to be clear these aren't cumulative speeds, but the speed from one sectional to the next (taking into account the distance travelled - Trakus data - for each horse), so from the start to 2f, from 2f to 3f, etc.  As shown, winners are slowing all the way to the finish, runners behind all show the same fatigue late only more sharply.

```{r, echo=FALSE, warning=FALSE, message=FALSE, out.height="350px", out.width="500px", fig.align='center'}
sspds <- select(df, pos, ss2f:ss6f)
y <- melt(sspds, id.vars = "pos")
y$variable <- gsub(pattern = "ss", replacement = "", x = y$variable)
y$variable <- gsub(pattern = "_", replacement = ".", x = y$variable)
y$variable <- factor(x = y$variable, levels = c("2f", "3f", "4f", "4.5f", "5f", "5.5f", "6f"))

ggplot(subset(y, pos ==1), aes(x = variable, y = value)) + 
    geom_boxplot() + 
    geom_jitter(position = position_jitter(width = .4), color=fill, alpha=.1) + 
    RcappeR::theme_rcapper() +
    labs(x = "Sectional", y = "Speed (ft per sec)", title="Sectional Speeds of Winners (6f dirt)")

wnrs_ind <- which.max(wnrs$ss2f)
all_ind <- which.max(df$ss2f)
```

The fastest opening quarter by a horse who went on to win was by **`r toupper(wnrs$horse[which.max(wnrs$ss2f)])`** on `r wnrs$date[which.max(wnrs$ss2f)]` in race `r wnrs$race[which.max(wnrs$ss2f)]`, recording a speed of **`r wnrs$ss2f[which.max(wnrs$ss2f)]`** feet per second.  `r if(wnrs$horse[wnrs_ind] != df$horse[all_ind]) { paste("There were horses who ran faster, ", toupper(df$horse[all_ind]), " ran the opening quarter in a speed of ", df$ss2f[all_ind], " ft per second.  ", if(wnrs$date_race[wnrs_ind] == df$date_race[all_ind]) { paste("Both horses were competing in the same ", toupper(df$race_type[all_ind]), " race", if(df$runup[all_ind] >= 50) { paste(", potentially benefitting from a runup of ", df$runup[all_ind], " feet.", sep = "") } else { paste0(".") }, sep = "") }, sep = "") } else if(df$date_race[all_ind] != wnrs$date_race[wnrs_ind]) { paste("The same horse, ", toupper(wnrs$horse[wnrs_ind]), ", recorded the fastest opening 2f in victory and in defeat, running at", df$ss2f[all_ind], "ft per sec when finishing ", df$pos[all_ind], " in race ", df$race[all_ind], " on ", df$date[all_ind], ".", sep = "") } else { paste("This opening quarter was the fastest speed by any horse, in victory or defeat.") }`

### Pace

Finishing Speeds also reflect the slowing to the finish shown in the Race Shape plot above, this will in part be due to the dirt surface (horses slow at the finish over longer trips also).  The par finishing speeds for `r dist`f winners are:

```{r echo=FALSE}
tmp <- wnrs %>%
    select(starts_with("fs")) %>%
    colMeans() %>%
    round(4)

names(tmp) = c("4f FinSpd", "3f FinSpd", "2f FinSpd", "1.5f FinSpd", "1f FinSpd", "0.5f FinSpd")
tmp
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
tmp <- mutate(df, Winner = ifelse(pos == 1, "Winners", "The Rest"))
tmp2 <- tmp %>%
    group_by(Winner) %>%
    summarise(c2f = round(mean(c2f, na.rm = TRUE),2),
              fs4f = round(mean(fs4f, na.rm = TRUE), 4))

win2f <- subset(tmp2, Winner == "Winners")$c2f
rest2f <- subset(tmp2, Winner != "Winners")$c2f
win_rest_diff <- abs(win2f - rest2f)

ldr2f <- round(mean(subset(df, p2f == 1)$c2f, na.rm = TRUE), 2)
```

The plots below show the distribution of 4f Finishing Speeds for winners compared to "The Rest", and the opening 2f sectionals for winners compared to "The Rest".  The distributions are similar, not surprising to see "The Rest" with a fatter distribution in both plots, perhaps most interesting is that the winners not only record higher Finishing Speeds than "The Rest", but they are also faster over the opening 2f.  Winners, on average, run the opening 2f in **`r win2f`**, compared to "The Rest", who run the opening 2f in **`r rest2f`**, a difference of **`r win_rest_diff`** seconds.  For reference, the horses leading their race after 2f run the opening 2f in an average time of **`r ldr2f`**.

```{r fig.show='hold', echo=FALSE, warning=FALSE, message=FALSE, out.height="300px", out.width="450px", fig.align='default'}
ggplot(tmp, aes(x = fs4f, y = ..density.. / 100, fill = Winner)) +
    geom_density(alpha = .5) +
    scale_fill_manual(values = c("Winners" = fill, "The Rest" = "#d8d8d8")) +
    RcappeR::theme_rcapper() +
    labs(x = "Finishing Speed", y = "", title = "4f Finishing Speed density plot (6f dirt)")

ggplot(tmp, aes(x = c2f, y = ..density.. / 100, fill = Winner)) +
    geom_density(alpha = .5) +
    scale_fill_manual(values = c("Winners" = fill, "The Rest" = "#d8d8d8")) +
    RcappeR::theme_rcapper() +
    labs(x = "Sectional (sec)", y = "", title = "Opening 2f Sectional density plot (6f dirt)")
```

Plotting the opening 2f sectional against the closing 4f Finishing Speeds the difference becomes a little more clear.  The cloud of points is crowded but winners are, for the majority, in the upper left part of the cloud.  

```{r echo=FALSE, warning=FALSE, message=FALSE, out.height="400px", out.width="550px", fig.align='center'}
ggplot(tmp, aes(x = c2f, y = fs4f, col = Winner)) + 
    geom_point(alpha = .5) + 
    scale_color_manual(values = c("Winners" = fill, "The Rest" = "#d8d8d8")) + 
    RcappeR::theme_rcapper() +
    labs(x = "Sectional (sec)", y = "Finishing Speed", title = "Opening 2f vs 4f Finishing Speed (6f dirt)")

fastestfs <- which.max(wnrs$fs4f)
fastestop <- which.min(wnrs$c2f)
```

The winner with the highest 4f Finishing Speed was **`r toupper(wnrs$horse[fastestfs])`** when winning a **`r wnrs$race_type[fastestfs]`** (race `r wnrs$race[fastestfs]` on `r wnrs$date[fastestfs]`), after an opening quarter of **`r wnrs$c2f[fastestfs]`** `r toupper(wnrs$horse[fastestfs])` recorded a Finishing Speed of **`r wnrs$fs4f[fastestfs]`**.  The winner with the fastest opening quarter was **`r toupper(wnrs$horse[fastestop])`** when winning a **`r wnrs$race_type[fastestop]`** (race `r wnrs$race[fastestop]` on `r wnrs$date[fastestop]`), recording a time of **`r wnrs$c2f[fastestop]`** over the opening quarter and a 4f Finishing Speed of **`r wnrs$fs4f[fastestop]`**.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
mongo.destroy(mongo)
rm(list=ls())
```