---
output: html_fragment
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
## This code unfortunately can't be run by others.
library(tomhes)
library(dplyr, warn.conflicts = FALSE, quietly = TRUE)
library(ggplot2)
library(reshape)

dist <- 7.5

mongo <- mongo.create()
if(mongo.is.connected(mongo)) {
    buf <- mongo.bson.buffer.create()
    mongo.bson.buffer.append(buf, name = "details.dist", value = dist)
    query <- mongo.bson.from.buffer(buf = buf)
    count <- mongo.count(mongo = mongo, ns = "gulf.races", query = query)
    cursor <- mongo.find(mongo = mongo, ns = "gulf.races", query = query)
    output <- vector("list", count)
    i <- 1
    while(mongo.cursor.next(cursor)) {
        output[[i]] <- mongo.bson.to.list(mongo.cursor.value(cursor))
        i <- i + 1
    }
    df <- gulf_races_df(races = output)
}
rm(buf, count, cursor, i, output, query)
fill = "#D9220F"
dirt <- subset(df, surf == "dirt")
turf <- subset(df, surf == "turf")
wnrs <- subset(dirt, pos == 1)
```
_Last Updated `r Sys.Date()`_

# Dirt

This is the track profile for all `r dist` furlong dirt races at Gulfstream Park, there are **`r length(unique(dirt$date_race))`** races in the dataset, contested by **`r length(dirt$date)`** runners.  The oldest race in the dataset was run on **`r min(as.Date(dirt$date, format = "%d/%m/%y"))`** while the most recent was run on **`r max(as.Date(dirt$date, format = "%d/%m/%y"))`**.

As there are only `r length(unique(dirt$date_race))` races in the dataset there is no profile.

```{r echo=FALSE, warning=FALSE, message=FALSE}
rm(dirt)
wnrs <- subset(turf, pos == 1)
```

# Turf

This is the track profile for all `r dist` furlong turf races at Gulfstream Park, there are **`r length(unique(turf$date_race))`** races in the dataset, contested by **`r length(turf$date)`** runners.  The oldest race in the dataset was run on **`r min(as.Date(turf$date, format = "%d/%m/%y"))`** while the most recent was run on **`r max(as.Date(turf$date, format = "%d/%m/%y"))`**.

### Summary

```{r echo=FALSE, warning=FALSE, message=FALSE}
racetypes <- turf %>%
    filter(pos == 1) %>%
    group_by(race_type) %>%
    summarise(n = n()) %>%
    arrange(desc(n))

racetypes$race_type <- factor(racetypes$race_type, levels = racetypes$race_type)

fieldsize <- turf %>%
    group_by(date_race) %>%
    summarise(field = n()) %>%
    group_by(field) %>%
    summarise(count = n())

fieldsize$n <- fieldsize$field * fieldsize$count
fieldsize$field <- factor(fieldsize$field, levels = fieldsize$field)
```

A few quick and dirty summary plots highlighting the different race types and field sizes.  **`r toupper(paste(racetypes$race_type[which.max(racetypes$n)]))`** races are the most common over the `r dist` furlong course, with **`r racetypes$n[which.max(racetypes$n)]`** races.  The average field size for `r dist` furlong races is **`r round(sum(fieldsize$n) / sum(fieldsize$count), 2)`**.

<div class="row">
<div class="col-sm-6">
```{r echo=FALSE, warning=FALSE, message=FALSE, out.height="300px", out.width="500px", fig.align='center'}
ggplot(racetypes, aes(x = race_type, y = n)) +
    geom_bar(fill = "#3D1995", stat = "identity") +
    RcappeR::theme_rcapper() +
    labs(x = "Race Type", y = "Count", title = "Counts per Race Type (7.5f Turf)") +
    theme(axis.text.x=element_text(angle=45, vjust=1, hjust=1, size = rel(1)))
```
</div>
<div class="col-sm-6">
```{r echo=FALSE, warning=FALSE, message=FALSE, out.height="300px", out.width="500px", fig.align='center'}
ggplot(fieldsize, aes(x = field, y = count)) +
    geom_bar(fill = "#3D1995", stat = "identity") +
    RcappeR::theme_rcapper() +
    labs(x = "No. of Runners", y = "Count", title = "Field Size (7.5f Turf)")

rm(racetypes, fieldsize)
```
</div>
</div>

### Draw

```{r echo=FALSE, warning=FALSE, message=FALSE}
pctbtn <- df %>%
    group_by(gate) %>%
    mutate(rnrs = n()) %>%
    ungroup() %>%
    filter(rnrs >= 5) %>%
    group_by(date_race) %>%
    mutate(rnrs_btn = RcappeR::rnrs_btn(pos)) %>%
    ungroup() %>%
    group_by(gate) %>%
    summarise(pct_btn = mean(rnrs_btn, na.rm = TRUE),
              mean_pos = mean(pos, na.rm = T))

pctbtn$gate <- factor(pctbtn$gate, levels=pctbtn$gate)

morethan55 <- pctbtn$gate[which(pctbtn$pct_btn >= .55)]

drawdist <- df %>%
    group_by(gate) %>%
    summarise(n = n(),
              mean_dist = mean(d7_5f, na.rm = TRUE))

drawdist <- filter(drawdist, n >= 5)

lowgate <- drawdist$gate[which.min(drawdist$mean_dist)]
lowdist <- round(subset(drawdist, gate == lowgate)$mean_dist, 2)
highgate <- drawdist$gate[which.max(drawdist$mean_dist)]
highdist <- round(subset(drawdist, gate == highgate)$mean_dist, 2)
```

The `r dist` furlong course is run round two turns, the first turn starts almost immediately after horse jump from the starting gates, and continues for roughly 2 furlongs, the second turn ends roughly 1f from the finish line.  Counting winners is the typical way to assess any draw bias, but winners ignore a large number of runners, and if there was any such bias, it should be seen across all runners, so percentage of runners beaten is to be used (gates must have a minimum of 5 runs).  A fair draw would expect to see each draw beat 50% of their rivals`r if(length(morethan55) > 0) { paste(", runners draw in", if(length(morethan55) >= 2) { paste("gates") } else { paste("gate") }) } else { paste(".") }` `r if(length(morethan55) > 0) { paste(morethan55, collapse = ", ") }` `r if(length(morethan55) > 0) { paste(" beat more than 55% of their rivals.") }`

The draw will likely have an impact on the distance horses travel, especially as the first turn is such a short distance from the start.  Horses drawn wide (high) will likely be forced wide.  `r if(drawdist$gate[which.min(drawdist$mean_dist)] == 1) { paste("Unsurprisingly, horses") } else { paste("Horses") }` drawn in gate **`r lowgate`** travel the least distance, with an average trip of **`r lowdist`** feet, while horses draw in gate **`r highgate`** travel an average distance of **`r highdist`**, a difference of **`r highdist - lowdist`** feet (gates must have a minimum of 5 runs used).  The plot below shows boxplots of distance raced by horses in each starting gate.

<div class="row">
<div class="col-sm-6">
```{r echo=FALSE, warning=FALSE, message=FALSE, out.height="300px", out.width="500px", fig.align='center'}
ggplot(data = pctbtn, aes(x = gate, y = pct_btn)) +
    geom_bar(fill = "#0BA42D", stat = "identity") +
    RcappeR::theme_rcapper() +
    labs(title = "Percentage of Rivals Beaten (7.5f Turf)", y = "")
```
</div>
<div class="col-sm-6">
```{r echo=FALSE, warning=FALSE, message=FALSE, out.height="300px", out.width="500px", fig.align='center'}
tmp <- df %>%
    group_by(gate) %>%
    mutate(runs = n()) %>%
    ungroup() %>%
    filter(runs >= 5)

ggplot(tmp, aes(x = factor(gate), y = d7_5f)) +
    geom_boxplot(fill = "#0BA42D") +
    RcappeR::theme_rcapper() +
    labs(x = "Gate", y = "Distance (ft)", title = "Distance Raced from each Starting Gate (7.5f Turf)")

rm(tmp, lowdist, lowgate, highdist, highgate)
```
</div>
</div>

### Race Shape

```{r, echo=FALSE, warning=FALSE, message=FALSE}
sspds <- select(wnrs, pos, ss2f:ss7_5f)
y <- melt(sspds, id.vars = "pos")
y$variable <- gsub(pattern = "ss", replacement = "", x = y$variable)
y$variable <- gsub(pattern = "_", replacement = ".", x = y$variable)
y$variable <- factor(x = y$variable, levels = c("2f", "4f", "5f", "6f", "6.5f", "7f", "7.5f"))

wnrs_ind <- which.max(wnrs$ss2f)
all_ind <- which.max(turf$ss2f)
```

<div class="row">
<div class="col-sm-6">

Race shape on turf courses isn't quite as predictable as on dirt courses, as the surface allows horses who started slowly to finish faster (you will see the difference in profiles for 8f and 8.5f races on dirt and turf).  The plot below shows the sectional speeds of winners, to be clear these aren't cumulative speeds, but the speed from one sectional to the next (taking into account the distance travelled - Trakus data - for each horse), so from the start to 2f, from 2f to 3f, etc.  Winners are still typically running slower over the final 0.5f but the fall off isn't as sharp as seen on dirt, and the middle part of the race from 2f to ~6.5f sees runners showing consistent speed.

The fastest opening quarter by a horse who went on to win was by **`r toupper(wnrs$horse[which.max(wnrs$ss2f)])`** on `r wnrs$date[which.max(wnrs$ss2f)]` in race `r wnrs$race[which.max(wnrs$ss2f)]`, recording a speed of **`r wnrs$ss2f[which.max(wnrs$ss2f)]`** feet per second.  `r if(wnrs$horse[wnrs_ind] != turf$horse[all_ind]) { paste("There were horses who ran faster, ", toupper(turf$horse[all_ind]), " ran the opening quarter in a speed of ", turf$ss2f[all_ind], " ft per second.  ", if(wnrs$date_race[wnrs_ind] == turf$date_race[all_ind]) { paste("Both horses were competing in the same ", toupper(turf$race_type[all_ind]), " race", if(turf$runup[all_ind] >= 50) { paste(", potentially benefitting from a runup of ", turf$runup[all_ind], " feet.", sep = "") } else { paste0(".") }, sep = "") }, sep = "") } else if(turf$date_race[all_ind] != wnrs$date_race[wnrs_ind]) { paste("The same horse, ", toupper(wnrs$horse[wnrs_ind]), ", recorded the fastest opening 2f in victory and in defeat, running at", turf$ss2f[all_ind], "ft per sec when finishing ", turf$pos[all_ind], " in race ", turf$race[all_ind], " on ", turf$date[all_ind], ".", sep = "") } else { paste("This opening quarter was the fastest speed by any horse, in victory or defeat.") }`

</div>
<div class="col-sm-6">
```{r, echo=FALSE, warning=FALSE, message=FALSE, out.height="300px", out.width="500px", fig.align='center'}
ggplot(subset(y, pos ==1), aes(x = variable, y = value)) +
    geom_boxplot() +
    geom_jitter(position = position_jitter(width = .4), color=fill, alpha=.25) +
    RcappeR::theme_rcapper() +
    labs(x = "Sectional", y = "Speed (ft per sec)", title="Sectional Speeds of Winners (7.5f Turf)")
```
</div>
</div>

### Pace

Finishing Speeds also reflect the consistent speed throughout the race, only falling off in the last 0.5f.

```{r echo=FALSE}
tmp <- wnrs %>%
    select(starts_with("fs")) %>%
    colMeans() %>%
    round(4)

names(tmp) = c("5.5f FinSpd", "3.5f FinSpd", "2.5f FinSpd", "1.5f FinSpd", "1f FinSpd", "0.5f FinSpd")
tmp
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
tmp <- mutate(turf, Winner = ifelse(pos == 1, "Winners", "The Rest"))
tmp2 <- tmp %>%
    group_by(Winner) %>%
    summarise(c4f = round(mean(c4f, na.rm = TRUE),2),
              fs3.5f = round(mean(fs3_5f, na.rm = TRUE), 4))

win2f <- subset(tmp2, Winner == "Winners")$c4f
rest2f <- subset(tmp2, Winner != "Winners")$c4f
win_rest_diff <- abs(win2f - rest2f)

ldr2f <- round(mean(subset(turf, p4f == 1)$c4f, na.rm = TRUE), 2)
```

The plots below show the distribution of 5f Finishing Speeds for winners compared to "The Rest", and the opening 2f sectionals for winners compared to "The Rest".  The distributions are similar, not surprising to see "The Rest" with a fatter distribution in both plots, perhaps most interesting is that the winners not only record higher Finishing Speeds than "The Rest", but they are also faster over the opening 2f.  Winners, on average, run the opening 2f in **`r win2f`**, compared to "The Rest", who run the opening 2f in **`r rest2f`**, a difference of **`r win_rest_diff`** seconds.  For reference, the horses leading their race after 2f run the opening 2f in an average time of **`r ldr2f`**.

<div class="row">
<div class="col-sm-6">
```{r echo=FALSE, warning=FALSE, message=FALSE, out.height="300px", out.width="450px", fig.align='center'}
ggplot(tmp, aes(x = fs3_5f, y = ..density.., fill = Winner)) +
    geom_density(alpha = .5) +
    scale_fill_manual(values = c("Winners" = fill, "The Rest" = "#d8d8d8")) +
    RcappeR::theme_rcapper() +
    labs(x = "Finishing Speed", y = "", title = "3.5f Finishing Speed density plot (7.5f Turf)")
```
</div>
<div class="col-sm-6">
```{r echo=FALSE, warning=FALSE, message=FALSE, out.height="300px", out.width="450px", fig.align='center'}
ggplot(tmp, aes(x = c4f, y = ..density.., fill = Winner)) +
    geom_density(alpha = .5) +
    scale_fill_manual(values = c("Winners" = fill, "The Rest" = "#d8d8d8")) +
    RcappeR::theme_rcapper() +
    labs(x = "Sectional (sec)", y = "", title = "Opening 4f Sectional density plot (7.5f Turf)")
```
</div>
</div>

```{r echo=FALSE, warning=FALSE, message=FALSE}
fastestfs <- which.max(wnrs$fs3_5f)
fastestop <- which.min(wnrs$c4f)
```
<div class="row">
<div class="col-sm-6">
Plotting the opening 4f sectional against the closing 3.5f Finishing Speeds the difference becomes a little more clear.  Runners aren't necessarily punished if they start slowly as in sprint, or dirt, races, the kinder surface means horses are capable of finishing much faster and making up ground on horses who started faster.

The winner with the highest 3.5f Finishing Speed was **`r toupper(wnrs$horse[fastestfs])`** when winning a **`r wnrs$race_type[fastestfs]`** (race `r wnrs$race[fastestfs]` on `r wnrs$date[fastestfs]`), after an opening 4f of **`r wnrs$c4f[fastestfs]`** `r toupper(wnrs$horse[fastestfs])` recorded a Finishing Speed of **`r wnrs$fs3_5f[fastestfs]`**.  The winner with the fastest opening quarter was **`r toupper(wnrs$horse[fastestop])`** when winning a **`r wnrs$race_type[fastestop]`** (race `r wnrs$race[fastestop]` on `r wnrs$date[fastestop]`), recording a time of **`r wnrs$c4f[fastestop]`** over the opening half mile and a 3.5f Finishing Speed of **`r wnrs$fs3_5f[fastestop]`**.

</div>
<div class="col-sm-6">

```{r echo=FALSE, warning=FALSE, message=FALSE, out.height="400px", out.width="550px", fig.align='center'}
ggplot(tmp, aes(x = c4f, y = fs3_5f, col = Winner)) +
    geom_point(alpha = .5) +
    scale_color_manual(values = c("Winners" = fill, "The Rest" = "#d8d8d8")) +
    RcappeR::theme_rcapper() +
    labs(x = "Sectional (sec)", y = "Finishing Speed", title = "Opening 4f vs 3.5f Finishing Speed (7.5f Turf)")
```
</div>
</div>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
y <- mongo.destroy(mongo)
rm(list=ls())
```
