---
output: html_fragment
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
## This code unfortunately can't be run by others.
library(tomhes)
library(dplyr, warn.conflicts = FALSE, quietly = TRUE)
library(ggplot2)
library(reshape)

mongo <- mongo.create()
if(mongo.is.connected(mongo)) {
    buf <- mongo.bson.buffer.create()
    mongo.bson.buffer.append(buf, name = "details.dist", value = 8.5)
    query <- mongo.bson.from.buffer(buf = buf)
    count <- mongo.count(mongo = mongo, ns = "gulf.races", query = query)
    cursor <- mongo.find(mongo = mongo, ns = "gulf.races", query = query)
    output <- vector("list", count)
    i <- 1
    while(mongo.cursor.next(cursor)) {
        output[[i]] <- mongo.bson.to.list(mongo.cursor.value(cursor))
        i <- i + 1
    }
    df <- gulf_races_df(races = output)
}
rm(buf, count, cursor, i, output, query)
fill = "#D9220F"
dirt <- filter(df, surf == "dirt")
dirtwnrs <- filter(dirt, pos == 1)
turf <- filter(df, surf == "turf")
turfwnrs <- filter(turf, pos == 1)
```
_Last Updated `r Sys.Date()`_

# Dirt

This is the track profile for all `r paste(unique(dirt$dist), collapse = ", ")` furlong `r paste(unique(dirt$surf), collapse = ", ")` races at Gulfstream Park, there are **`r length(unique(dirt$date_race))`** races in the dataset, contested by **`r length(dirt$date)`** runners.  The oldest race in the dataset was run on **`r min(as.Date(dirt$date, format = "%d/%m/%y"))`** while the most recent was run on **`r max(as.Date(dirt$date, format = "%d/%m/%y"))`**.

# Turf

This is the track profile for all `r paste(unique(turf$dist), collapse = ", ")` furlong `r paste(unique(turf$surf), collapse = ", ")` races at Gulfstream Park, there are **`r length(unique(turf$date_race))`** races in the dataset, contested by **`r length(turf$date)`** runners.  The oldest race in the dataset was run on **`r min(as.Date(turf$date, format = "%d/%m/%y"))`** while the most recent was run on **`r max(as.Date(turf$date, format = "%d/%m/%y"))`**.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
mongo.destroy(mongo)
rm(list=ls())
```